Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-10-13 astronum_2017/astronum-amrex.tex

\documentclass[letterpaper]{jpconf}

@@

@@

No single algorithm meets all of the demands imposed by these events.
Instead, we advance our understanding of these events by piecing
together simulations of different phases of the evolution from
different codes.  Here we discuss our simulation codes, \maestro\ and
\castro\ designed to perform three-dimensional models of the early
subsonic evolution leading to runaway and the subsequent expolosion.
Together this suite of codes allows us to address many problems in
stellar and nuclear astrophysics.  We describe some of the design
details, the current architecture of the code, and some applications
below.

@@

Our interests are thermonuclear explosions, including Type Ia
supernovae, X-ray bursts, and novae.  The basic ingredients for these
events are thermonuclear energy release and a degenerate equation of
state that allows a runaway to build without a pressure response.
Most of the current models for these events are characterized by a
long timescale ``simmmering'' phase where reactions heat the star or
layer and drive convection.  Eventually, reactions become vigorous
enough that a runaway takes place, perhaps with an accompanying
burning front that spreads through the star.

@@

There are open questions in all of these scenarios which can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? and can we
avoid an accretion-induced collapse to a neutron star.  For many of
these, we have the question of whether it is possible to accurately
model the ignition of a detonation with the spatial resolution we can
attain?  These are some of the questions we seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in C++ with the
computational kernels written in Fortran\footnote{Currently
  \maestro\ is using an older pure-Fortran interface, but this will be
  ported to the C++ AMReX this coming year.}---this allows us to take
advantage of the strengths of both languages.AMReX supports subcycling
in time, which we use in \castro.

@@

The state is advanced using a second-order accurate projection method.
Fluid quantities are advected using an unsplit Godunov method, with
reactions incorporated via operator splitting.  The provisional
velocities are then projected onto the space that satisfies the
divergence constraint.  The projections involve elliptic solves,
solving a variable-coefficient Poisson problem, computed using a
multigrid algorithm.  Low speed methods have seen a lot of recent
development~\cite{kleinpauluis,vasil:2013}, which has been
incorporated into \maestro.  The next development effort for
\maestro\ will be to allow for moderate rotation of the star.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity using geometric
multigrid.  The main hydrodynamics scheme in \castro\ is an unsplit
piecewise parabolic method.  Recent developments have improved
conservation with gravity sources and rotation~\cite{wdmergerI}.  The
radiation solver in \castro\ uses the flux-limited diffusion
approximation for gray or multigroup radiation.  The integration
algorithm on the grid hierarchy is a recursive procedure in which
coarse grids are advanced in time, fine grids are advanced multiple
steps to reach the same time as the coarse grids and the data at
different levels are then synchronized. The synchronization for
self-gravity is similar to the algorithm introduced by
\cite{miniati-colella}.

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ astrophysics github
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that that rates and integration strategy are decoupled, allowing us to
change the integration strategy for a set of rates.  They are also
written to be threadsafe and with GPUs in mind (more on that
below).  The goal of \starkiller\ is to evolve the code so that it is
able to be used in a variety of nuclear astrophysics codes, not just
those discussed here.

@@

All of our simulation codes are open source and follow a fully open
development model---the development git repos are hosted on
github\footnote{\url{https://github.com/AMReX-Astro/}}, available for
anyone to see and contribute to using issues and pull-requests.
New changes are put into the {\tt development} branch in each repo.
Nightly regression testing is used to ensure that no new bugs were
introduced.  Once a month, {\tt development} is merged into {\tt
  master}.
{\em All source files, model files, input parameters, etc.\ for any
published science results are also available in the code repos.}  When
feasible, the git hashes for the published results are included in paper
acknowledgements.

@@

Figure~\ref{fig:scaling} shows strong scaling for \maestro\ on the 3-d
XRB problem.  We ran on both OLCF titan and NERSC edison.  We see that
the code scales well to $\mathcal{O}(10^4)$ processors.  The main
limitation to the scaling at the moment are the multigrid solves used
to enforce the projection (in particular the nodal solver).
\maestro\ also does not yet take advantage of the tiling capabilities
offered in \amrex.  These will be addressed in the future as we port
\maestro\ to the C++ \amrex\ codebase to take advantage of ongoing
development.

@@

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex.  In \amrex, grids
consist of a number of boxes that enable domain decomposition for
parallelism.  An iterator loops over all of the boxes at the same
level of refinement and passes a data pointer into a Fortran kernel
function where the work is done. For the simulation state data that
resides in each box, we have modified the memory allocator so that it
can use a CUDA allocator (either device or managed memory).  The
domain iterator is configured to handle data motion to and from the
device, so that the compute kernels can operate on data which is
presumed to already be there, and the computation is decoupled from
the memory management. Computation on the data can then be performed
with OpenACC, CUDA Fortran, or (more recently) OpenMP 4.5. We have
also built CUDA compute support into \amrex\ so that a compute kernel
can be transparently operated on using CUDA Fortran without
substantially modifying the kernels (and we anticipate using a similar
strategy to use OpenACC and/or OpenMP in the future). This helps
ensure that we can continue to maintain performance portability in our
simulation codes, by decoupling the physics algorithms from the
backend support used to implement them on various compute
architectures.\MarginPar{add CPU time for reference}

@@

Figure~\ref{fig:current-runs} shows some of our recent science
simulations.  The left panel is an image of convection in the helium
layer on a sub-Chandra white dwarf.  This is part of a study of the
early stages of the double detonation SNe Ia model.  Using \maestro,
we are able to model the convection in the He layer for many turnover
times and saw a range of outcomes depending on the mass of the white
dwarf and He layer, including a global He-nova-like behavior where the
entire layer runsway together and localized
burning~\cite{subchandra2}.

The middle figure shows convection in a H/He layer on a neutron star,
as a model of the early burning in an XRB.  This \maestro\ model was
the first 3D model of convection for this problem~\cite{xrb3}.  This
study showed that the convective field became fully turbulent,
achieving a Kolmogorov spectrum, and the overall dynamics was very
different than our earlier 2-d simulations.

The rightmost image is the coalesced remains of the merger of a
0.9~$M_\odot$ and 0.6~$M_\odot$ WD performed with \castro.  This used
the developments from \cite{wdmergerI}.  Our primary focus with this
suite of simulations is understanding the numerical sensitivity of
mergers and collisions on the burning that takes place.

@@

We described our suite of astrophysics codes built on the
\amrex\ framework.  These codes were developed to model problems in
stellar astrophysics spanning from low speed convection to explosive
burning.  A major theme of the codes is the open development model,
with the code development done on github and all problem files needed
to recreate any science results freely available.  Future development
efforts for \maestro\ include higher-order hydrodynamics and
time-integration and rotation.  For \castro, we are investigating
strong coupling between hydrodynamics and reactions, new solvers
(including MHD), and finishing the GPU port.

>>>>

\documentclass[a4paper]{jpconf}

@@

\newcommand{\cpp}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf +}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\tiny\bf +}}

@@

No single algorithm meets all of the demands imposed by these events.
Instead, we advance our understanding of these events by piecing
together simulations of different phases of the evolution from
different codes.  Here we discuss our simulation codes, \maestro\ and
\castro, designed to perform three-dimensional models of the early
subsonic evolution leading to runaway and the subsequent expolosion.
Together this suite of codes allows us to address many problems in
stellar and nuclear astrophysics.  We describe some of the design
details, the current architecture of the code, and some applications
below.

@@

Our interests are thermonuclear explosions, including Type Ia
supernovae (SNe Ia), X-ray bursts (XRBs), and novae.  The basic
ingredients for these events are thermonuclear energy release and a
degenerate equation of state that allows a runaway to build without a
pressure response.  Most of the current models for these events are
characterized by a long timescale ``simmmering'' phase where reactions
heat the star or layer and drive convection.  Eventually, reactions
become vigorous enough that a runaway takes place, perhaps with an
accompanying burning front that spreads through the star.

@@

There are open questions in all of these scenarios which can be
addressed through simulation.  For the Chandra and sub-Chandra models,
what is the distribution (spatial and temporal) of the hotspots setup
by turbulent convection that give rise to burning fronts?  A
longstanding question with the Chandra model is, can a deflagration
transition to a detonation during the burning front propagation
through the star.  For the sub-Chandra double detonation model, is it
possible to create a detonation in the thin surface He layer?  For
double degenerates, does the burning take place promptly? and can we
avoid an accretion-induced collapse to a neutron star.  For many of
these, we also need to address numerical issues, such as whether it is
possible to accurately model the ignition of a detonation with the
spatial resolution we can attain?  These are some of the questions we
seek to answer.

@@

Our suite of application codes is built on the
\amrex\ block-structured adaptive mesh refinement (AMR) library.  The
basic programming model has \amrex\ managing the grid data-structures
and parallel communications, and it calls the computational kernels on
a patch-by-patch basis.  The core library is written in \cpp\ with the
computational kernels written in Fortran\footnote{Currently
  \maestro\ is using an older pure-Fortran interface, but this will be
  ported to the \cpp\ AMReX this coming year.}---this allows us to take
advantage of the strengths of both languages.AMReX supports subcycling
in time, which we use in \castro.

@@

The state is advanced using a second-order accurate projection method.
Fluid quantities are advected using an unsplit Godunov method, with
reactions incorporated via operator splitting.  The provisional
velocities are then projected onto the space that satisfies the
divergence constraint.  The projections involve elliptic solves,
solving a variable-coefficient Poisson problem, computed using a
multigrid algorithm.  Low speed methods have seen a lot of recent
development~\cite{kleinpauluis,vasil:2013}, which has been
incorporated into \maestro.

@@

\castro~\cite{castro,castroII,castroIII} is a fully-compressible
radiation hydrodynamics code that supports arbitrary equations of
state, nuclear reaction networks, and Poisson gravity using geometric
multigrid.  The main hydrodynamics scheme in \castro\ is an unsplit
piecewise parabolic method.  The radiation solver in \castro\ uses the
flux-limited diffusion approximation for gray or multigroup radiation.
The integration algorithm on the grid hierarchy is a recursive
procedure in which coarse grids are advanced in time, fine grids are
advanced multiple steps to reach the same time as the coarse grids and
the data at different levels are then synchronized. The
synchronization for self-gravity is similar to the algorithm
introduced by \cite{miniati-colella}.  Recent developments in
\castro\ include a deferred-correction method of coupling
hydrodynamics and reactions, a conservative gravity and rotation
source formulation~\cite{wdmergerI}, and retry mechanism to redo a
step based on criteria evaluated during the integration.\MarginPar{more new stuff?}

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ astrophysics github
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that that rates and integration strategy are decoupled, allowing us to
change the integration strategy for a set of rates.  They are also
written to be threadsafe and with GPUs in mind (more on that below).
The goal of \starkiller\ is to create a set of community microphysics
solvers that can be used in a variety of nuclear astrophysics codes,
not just those discussed here.

@@

All of our simulation codes are open source and follow a fully open
development model---the development git repos are hosted on
github\footnote{\url{https://github.com/AMReX-Astro/}}, available for
anyone to see and contribute to using issues and pull-requests.
Additionally, we have mailing lists for discussions and asking for
help.  Several branches are used in our workflow.  New changes are put
into the {\tt development} branch in each repo.  Nightly regression
testing is used to ensure that no new bugs were introduced.  Once a
month, {\tt development} is merged into {\tt master}.  Finally, {\em all source
  files, model files, input parameters, etc.\ for any published
  science results are also available in the code repos.}  When
feasible, the git hashes for the published results are included in
paper acknowledgements.

@@

Figure~\ref{fig:maestro_scaling} shows strong scaling for \maestro\ on the 3-d
XRB problem.  This is a typical \maestro\ application~\cite{xrb3}, where burning
can be a significant part of the overall evolution.  We ran on both
OLCF titan and NERSC edison.  We see that the code scales well to
$\mathcal{O}(10^4)$ processors.  The main limitation to the scaling at
the moment are the multigrid solves used to enforce the projection (in
particular the nodal solver).  \maestro\ also does not yet take
advantage of the tiling capabilities offered in \amrex.  These will be
addressed in the future as we port \maestro\ to the
\cpp\ \amrex\ codebase to take advantage of ongoing development.

@@

Our latest focus has been on GPU ports of our application codes.  A
small proxy app, \starlord, was created from \castro\ with just the
hydrodynamics and stellar equation of state.  It uses a simple
method-of-lines formulation of the hydrodynamics and advects 13
nuclear species in addition to the hydrodynamics.  To offload work on
GPUs, GPU support was added directly into \amrex\footnote{This support
  is currently on a feature branch in the git repo, awaiting merge
  into {\tt development}.}.  In \amrex, grids consist of a number of
boxes that enable domain decomposition for parallelism.  An iterator
loops over all of the boxes at the same level of refinement and passes
a data pointer into a Fortran kernel function where the work is
done. For the simulation state data that resides in each box, we have
modified the memory allocator so that it can use a CUDA allocator
(either device or managed memory).  The domain iterator is configured
to handle data motion to and from the device, so that the compute
kernels can operate on data which is presumed to already be there, and
the computation is decoupled from the memory management. Computation
on the data can then be performed with OpenACC, CUDA Fortran, or (more
recently) OpenMP 4.5. We have also built CUDA compute support into
\amrex\ so that a compute kernel can be transparently operated on
using CUDA Fortran without substantially modifying the kernels (and we
anticipate using a similar strategy to use OpenACC and/or OpenMP in
the future). This helps ensure that we can continue to maintain
performance portability in our simulation codes, by decoupling the
physics algorithms from the backend support used to implement them on
various compute architectures.\MarginPar{add CPU time for reference}

@@

Figure~\ref{fig:current-runs} shows some of our recent science
simulations.  The left panel is an image of convection in the helium
layer on a sub-Chandra white dwarf.  This is part of a study of the
early stages of the double detonation SNe Ia model.  Using \maestro,
we are able to model the convection in the He layer for many turnover
times and saw a range of outcomes depending on the mass of the white
dwarf and He layer, including a global He-nova-like behavior where the
entire layer runsway together and localized
burning~\cite{subchandra2}.  Work continues to characterize the ignition.

The middle figure shows convection in a H/He layer on a neutron star,
as a model of the early burning in an XRB.  This \maestro\ model was
the first 3D model of convection for this problem~\cite{xrb3}.  This
study showed that the convective field became fully turbulent,
achieving a Kolmogorov spectrum, and the overall dynamics was very
different than our earlier 2-d simulations.  This calculation acts as 
a bridge to our next set of studies that will look at larger scales.

The rightmost image is the coalesced remains of the merger of a
0.9~$M_\odot$ and 0.6~$M_\odot$ WD performed with \castro.  This used
the developments from \cite{wdmergerI}.  Our primary focus with this
suite of simulations is understanding the numerical sensitivity of
mergers and collisions on the burning that takes place.  This work
is ongoing.

@@

We described our suite of astrophysics codes built on the
\amrex\ block-structured adaptive mesh refinement framework.  These
codes were developed to model problems in stellar astrophysics
spanning from low speed convection to explosive burning.  A major
theme of the codes is the open development model, with the code
development done on github and all problem files needed to recreate
any science results freely available.  Future development efforts for
\maestro\ include higher-order hydrodynamics and time-integration and
rotation.  For \castro, we are investigating stronger coupling between
hydrodynamics and reactions, new solvers (including MHD), and
finishing the GPU port.
